---
title: "Human Brain (DLPFC) 10x EDA & Seurat"
author: "Emily Ekstrum"
date: ""
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: cosmo
params:
  # gene expression
  h5_path: "/Users/emilyekstrum/Desktop/zhangLab/data/DLPFC_Data/Sample151673/151673_filtered_feature_bc_matrix.h5"

  # spatial coordinates
  spatial_csv: "/Users/emilyekstrum/Desktop/zhangLab/data/DLPFC_Data/Sample151673/DLPFC_spatial_3639cells_2cols.csv"

  # manual annotations
  annotation_csv: "/Users/emilyekstrum/Desktop/zhangLab/data/DLPFC_Data/Sample151673/Manual_annotation_3639cells.csv"

  # for seurat object
  assay_name: "RNA"
  min_cells: 3
  min_features: 200

  # filtering params
  filter: true
  nFeature_low: 200
  nFeature_high: 7500
  nCount_high: 60000
  percent_mt_high: 20

  # dim reduction params
  n_pcs: 30
  clustering_resolution: 0.5
  seed: 123
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
set.seed(params$seed)

`%||%` <- function(a, b) if (!is.null(a)) a else b
```

# Overview

This report reproduces the same Seurat pipeline you ran for the mouse brain data, but for the **human brain (DLPFC) dataset** stored as a 10x `.h5` matrix:

1. Load 10x counts (`Read10X_h5`)
2. Optionally merge spatial coordinates and manual annotations into `meta.data`
3. EDA/QC: library size, detected genes, mitochondrial fraction
4. Optional filtering
<!-- 5. Standard Seurat workflow: Normalize → HVFs → Scale → PCA → Neighbors → Clusters → UMAP -->
6. Spatial visualization using coordinate scatter plots (works even without histology images)

**Inputs**
- Counts: `r params$h5_path`
- Spatial coords (optional): `r params$spatial_csv`
- Manual annotations (optional): `r params$annotation_csv`

# Packages

```{r packages}
pkgs <- c("Seurat","Matrix","ggplot2","dplyr","tibble","patchwork","readr", "hdf5r")
to_install <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(to_install) > 0) install.packages(to_install, repos = "https://cloud.r-project.org")
lapply(pkgs, library, character.only = TRUE)
```

# Load counts from 10x H5

```{r load-h5}
h5_path <- params$h5_path
stopifnot(file.exists(h5_path))

counts <- Read10X_h5(h5_path)

# Read10X_h5 may return a matrix OR a named list of matrices (e.g., "Gene Expression")
if (is.list(counts)) {
  message("Read10X_h5 returned a list with names: ", paste(names(counts), collapse = ", "))
  # Prefer common key names
  key <- intersect(names(counts), c("Gene Expression","RNA","gene_expression","GeneExpression"))
  if (length(key) == 0) key <- names(counts)[1]
  counts <- counts[[key[1]]]
}

stopifnot(inherits(counts, c("dgCMatrix","matrix")))
dim(counts)
```

# Load optional metadata (spatial coords + manual annotations)

This section tries to merge metadata by cell barcode when possible. If no explicit barcode column is found, it falls back to matching by row order only when the row count matches the number of cells.

```{r load-meta}
load_csv_if_exists <- function(path) {
  if (is.null(path) || path == "" || is.na(path) || !file.exists(path)) return(NULL)
  df <- readr::read_csv(path, show_col_types = FALSE)
  as.data.frame(df)
}

spatial_df <- load_csv_if_exists(params$spatial_csv)
annot_df   <- load_csv_if_exists(params$annotation_csv)

if (!is.null(spatial_df)) {
  message("Loaded spatial CSV with dim: ", paste(dim(spatial_df), collapse=" x "))
  print(head(spatial_df, 3))
} else {
  message("No spatial CSV loaded (file missing or params$spatial_csv is blank).")
}

if (!is.null(annot_df)) {
  message("Loaded annotation CSV with dim: ", paste(dim(annot_df), collapse=" x "))
  print(head(annot_df, 3))
} else {
  message("No annotation CSV loaded (file missing or params$annotation_csv is blank).")
}
```

## Merge helper

```{r merge-helper}
guess_barcode_col <- function(df) {
  if (is.null(df)) return(NULL)
  nms <- tolower(names(df))
  candidates <- c("barcode","barcodes","cell","cell_id","cellid","spot","spot_id","id")
  hit <- candidates[candidates %in% nms]
  if (length(hit) > 0) return(names(df)[match(hit[1], nms)])
  NULL
}

merge_meta <- function(meta, df, cells) {
  if (is.null(df)) return(meta)

  bc_col <- guess_barcode_col(df)

  if (!is.null(bc_col)) {
    df[[bc_col]] <- as.character(df[[bc_col]])
    rownames(df) <- df[[bc_col]]
    df[[bc_col]] <- NULL
    df2 <- df[intersect(rownames(df), cells), , drop = FALSE]
    meta[rownames(df2), names(df2)] <- df2
    return(meta)
  }

  if (nrow(df) == length(cells)) {
    rownames(df) <- cells
    meta[cells, names(df)] <- df
    return(meta)
  }

  warning("Could not merge metadata: no obvious barcode column and row count did not match cell count.")
  return(meta)
}
```

# Create Seurat object

```{r create-seurat}
cells <- colnames(counts)
meta <- data.frame(row.names = cells)

meta <- merge_meta(meta, spatial_df, cells)
meta <- merge_meta(meta, annot_df,   cells)

seu <- CreateSeuratObject(
  counts = counts,
  assay = params$assay_name,
  min.cells = params$min_cells,
  min.features = params$min_features,
  meta.data = meta
)

seu
```

# EDA and QC

## Library size, detected genes

```{r qc-basic}
VlnPlot(seu, features = c("nFeature_RNA","nCount_RNA"), pt.size = 0.1, ncol = 2)
FeatureScatter(seu, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
```



## Mitochondrial fraction

```{r qc-mito}
mt_genes <- grep(params$mt_prefix, rownames(seu), value = TRUE)
if (length(mt_genes) > 0) {
  seu[["percent.mt"]] <- PercentageFeatureSet(seu, features = mt_genes)
  VlnPlot(seu, features = "percent.mt", pt.size = 0.1)
  FeatureScatter(seu, feature1 = "nCount_RNA", feature2 = "percent.mt")
} else {
  message("No mitochondrial genes matched pattern: ", params$mt_prefix)
}
```

# Optional filtering

```{r filtering}
if (isTRUE(params$filter)) {
  keep <- seu$nFeature_RNA >= params$nFeature_low &
          seu$nFeature_RNA <= params$nFeature_high &
          seu$nCount_RNA   <= params$nCount_high

  if ("percent.mt" %in% colnames(seu@meta.data)) {
    keep <- keep & (seu$percent.mt <= params$percent_mt_high)
  }

  message("Keeping ", sum(keep), " / ", length(keep), " cells after QC filters.")
  seu <- subset(seu, cells = colnames(seu)[keep])
}

seu
```

# Standard Seurat workflow (PCA → neighbors → clusters → UMAP)

```{r seurat-workflow}
seurat_sct <- SCTransform(
  seu,
  assay = "RNA",          # use the raw counts assay
  vst.flavor = "v2",  
  variable.features.n = 2000,
  verbose = FALSE
)

# make SCT the default assay for downstream PCA/UMAP
DefaultAssay(seurat_sct) <- "SCT"
seurat_sct <- RunPCA(seurat_sct, features = VariableFeatures(seurat_sct))

Reductions(seurat_sct)
ElbowPlot(seurat_sct, ndims = max(50, params$n_pcs))
```

```{r neighbors-cluster-umap}
pcs <- 1:params$n_pcs

seurat_sct <- FindNeighbors(seurat_sct, dims = pcs)
message("Graphs present: ", paste(names(seurat_sct@graphs), collapse = ", "))

seurat_sct <- FindClusters(seurat_sct, resolution = params$clustering_resolution)
seurat_sct <- RunUMAP(seurat_sct, dims = pcs)

DimPlot(seurat_sct, reduction = "umap", label = TRUE) + NoLegend()
```

# Spatial visualization (coordinate scatter plots)

This does **not** require histology images. It just uses whatever coordinate columns exist in `meta.data`.

## Detect coordinate columns

```{r coords-detect}
meta_cols <- colnames(seurat_sct@meta.data)
coord_candidates <- list(
  x = c("x","imagecol","col","pxl_col_in_fullres","array_col"),
  y = c("y","imagerow","row","pxl_row_in_fullres","array_row")
)

find_first_present <- function(cands, cols) {
  hit <- cands[cands %in% cols]
  if (length(hit) == 0) return(NULL)
  hit[1]
}

xcol <- find_first_present(coord_candidates$x, meta_cols)
ycol <- find_first_present(coord_candidates$y, meta_cols)

list(xcol = xcol, ycol = ycol)
```

## Plot clusters on tissue coordinates

```{r spatial-clusters}
if (!is.null(xcol) && !is.null(ycol)) {
  df <- data.frame(seurat_sct@meta.data, cluster = Idents(seurat_sct))
  ggplot(df, aes(x = .data[[xcol]], y = .data[[ycol]], color = cluster)) +
    geom_point(size = 0.8) +
    coord_fixed() +
    scale_y_reverse() +
    theme_classic() +
    labs(title = "Clusters in tissue coordinates", x = xcol, y = ycol)
} else {
  cat("No recognizable coordinate columns found in meta.data.\\n")
  cat("Available meta.data columns include:\\n")
  print(head(colnames(seurat_sct@meta.data), 30))
}
```

## Plot a marker gene on tissue coordinates

```{r spatial-gene}
if (!is.null(xcol) && !is.null(ycol)) {
  gene <- "MBP"  # change 
  if (gene %in% rownames(seurat_sct)) {
    df <- data.frame(seurat_sct@meta.data, expr = FetchData(seurat_sct, vars = gene)[,1])
    ggplot(df, aes(x = .data[[xcol]], y = .data[[ycol]], color = expr)) +
      geom_point(size = 0.8) +
      coord_fixed() +
      scale_y_reverse() +
      theme_classic() +
      labs(title = paste(gene, "expression in tissue coordinates"), x = xcol, y = ycol)
  } else {
    cat("Gene ", gene, " not found. Try another marker (e.g., SNAP25, AQP4, P2RY12, GFAP, MOBP).\\n")
  }
}
```

# Save outputs

```{r save}
out_dir <- "outputs_human"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

saveRDS(seu, file.path(out_dir, "human_dlpfc_seurat.rds"))

p_umap <- DimPlot(seu, reduction = "umap", label = TRUE) + NoLegend()
ggsave(filename = file.path(out_dir, "umap_clusters.png"), plot = p_umap, width = 7, height = 5, dpi = 300)

cat("Saved outputs to: ", out_dir, "\\n")
```

```{r session-info}
sessionInfo()
```
